How is experience transformed into memory?

### 2018-08-28 paper/main.tex
edits to results and figure captions

\begin{figure}[th!]
\centering
\includegraphics[width=1\textwidth]{figs/1_analysis_schematic_bw.pdf}
\caption{\small \textbf{Schematic of the analysis approach.} For each scene in the video, text descriptions were generated by hand. Three exemplary time points are displayed here.  Below the video descriptions are text samples from an example participant's verbal recall transcript.  We fit a topic model to the moment-by-moment video text descriptions and transformed participant's verbal recall transcripts using this same model. The bar charts display the resulting topic model weights for the video (in blue) and recall (in green) for three example topic dimensions.}
\label{fig:schematic}
\end{figure}

\subsubsection{Modeling the naturalistic stimulus and verbal recall}
We fit a topic model \citep{BleiEtal03} to hand-annotated text descriptions of scenes from the video. The text descriptions contained details of the scene such as the characters, location, and a short summary of the scene (see Fig.~\ref{fig:schematic} for example text). As depicted in Fig.~\ref{fig:model}a, the topic vectors are sparse and change slowly over time. Furthermore, there are clear transitions from one topic ``state'' to the next, possibly indexing scene transitions in the stimulus. To get a better handle on this temporal structure, we computed a timepoints (1976) by timepoints (1976) correlation matrix of the video model (Fig.~\ref{fig:model}c).  This correlation matrix reveals that the model has a strong, block-diagonal structure with very low off-diagonal correlation values, suggesting the representations of each block are unique and highly descriminable. We found that the ``narrative details'' (e.g. a few sentences summarizing the specific happenings of the scene) of the video were the most important feature driving this model structure (see Supp. Fig.~\ref{fig:feature-importance}a).

After watching the video, participants verbally recalled (in order) as much of the episode as they could.  We used the same topic model (fit with the text descriptions of the video) to transform participants' verbal recall transcripts. An example participant's (\#13) recall model is plotted in Fig.~\ref{fig:model}b. Notably, participant \#13's recall model appears visually similar to the video model. Similar to the video model analysis (Supp. Fig.~\ref{fig:feature-importance}a), We found that the ``narrative details'' feature was most important for driving the video/recall relationship (Supp. Fig.~\ref{fig:feature-importance}b).  Like the video model, topic vectors were sparse and changed gradually.  Then, we investigated the temporal structure of the recall matrices. From each participant's recall model, we computed a timepoint (68 to 294) by timepoints by timepoints (range: 68 to 294) correlation matrix. An example participant's (\#13) correlation matrix is plotted in Fig.~\ref{fig:model}d. Like the video model, each participant's recall correlation matrix exhibited a strong block-diagonal structure (Supp. Fig.~\ref{fig:corrmats}). Notably, this suggests that participants recounted the video in discriminable segments, likely related to the recall of specific events from the video.

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{figs/2_eventseg.pdf}
\caption{\small \textbf{Modelling naturalistic stimuli and recall.} A depiction of our analysis pipeline. For all plots, darker colors indicate greater values and the range of each plot is 0-1.  A). A timepoints (1976) by topics (100) matrix representing the video stimulus.  Each row represents the most likely mixture of topics for a given timepoint (i.e., topic weights). Each column represents a different topic. B). A timepoints (1976) by topics (100) matrix representing participant \#13's recall. C). A viewing-time (1976) by viewing-time (1976) correlation matrix representing the correlation of each moment of the video model with every other moment of the video model. The white boxes represent `events' recovered by a hidden Markov model. D). A recall-time (294 sentences) by recall-time (294) correlation matrix for participant \#13. E). An events (34) by topics (100) matrix where each row represents the average topic vector for each event in the video model.  F). An events (27) by topics (100) matrix where each row represents the average topic vector for each event in participant \#13's recall model. G. A recall events (27) by video events (34) correlation matrix for participant \#13. The cells with a yellow border identify the video event with the highest correlation to a given recall event. F. A group averaged recall events (34) by video events (34) correlation matrix.  The cells with yellow borders are the video event with the highest correlation to a given average recall event.}
\label{fig:model}
\end{figure}

\subsubsection{Segmenting the video and recall models into ``events''}
As described above, a striking feature of both the video and recall correlation matrices is a strong, block structure along the diagonal of the matrices (see Fig.~\ref{fig:model}c,d).  We hypothesized that this structure might arise from transient stability in the "theme" of both the video and of participants' memory for the video. We hypothesized that specific events described by a participant could be ``matched'' to specific video events by computing the similarity (i.e., correlation) between their topic vectors. In other words, the video event that is most similar to a particular recall event is the event that the participant is most likely describing. To test this idea, we segmented the video and recall models in $k$ events (i.e., states) using a hidden Markov model ~\citep{BaldEtal17}. Our algorithm determined 34 events for the video model and a range of values (range: 8-27; mean: 15.41; SD: 5.6) for the recall models (see Methods for details on choosing an optimal $k$ value).  The events discovered for the video model and a representative participant's recall model are highlighted as yellow rectangles outlining blocks along the diagonal of the correlation matrices (Fig.~\ref{fig:model}b,d).

Next, we created a video ``event model'' by averaging together neighboring topic vectors that the were classified to be in the same event, resulting in an events (34) by topics (100) matrix (Fig.~\ref{fig:model}e).  We performed the same procedure for the recall matrices (Fig.~\ref{fig:model}f). Then, we computed the correlation between video and recall event models, resulting in a video events (34) by recall events (8-27, depending on the participant) correlation matrix (Fig~\ref{fig:model}g for example, Supp. Fig~\ref{fig:matchmats} for all participants). These matrices represent the similarity (correlation) between each video event and each recall event (for each participant). To determine which video event a particular recall event was most likely describing, we computed the index of the video event with the highest correlation to the recall event (i.e., the argmax).  This is depicted in Fig.~\ref{fig:model}g as the cells highlighted with a yellow border. Notably, our algorithm suggests that the example participant recalled most of the video events and did so in order.

Then, we computed a group-averaged recall event model and video-recall event correlation matrix. For each participant (and each recall event), we sorted the recall event vectors (across all participants) according to the video event with the highest correlation. We then averaged the recall event vectors within each group. This yielded an average recall event vector for all but one (of 34) video events, since no participant remembered one of the events according to our model. Lastly, we computed an average recall event (34) by video event (34) correlation matrix, and highlighted the cell with the highest correlation value with a yellow border (Fig.~\ref{fig:model}h). Notably, this matrix displayed high correlation values along the diagonal and low correlations in the off-diagonal cells. This suggests that on average, participants were able to recapitulate the events in the episode in a specific and highly descriminable way.

\subsubsection{The shape of a naturalistic experience is preserved in recall}
Classic approaches to studying free recall in episodic memory (such as overall accuracy and serial position curve) focus on the quantity of information recalled \citep{Murd62a}. More recent metrics (such as lag-CRP, semantic and temporal clustering) go further to describe list-level memory organization \citep{Kaha96, PolyEtal09}. However, these approaches cannot (and were not designed to) capture the rich temporal dynamics inherent in naturalistic stimuli and associated recall of those experiences. In the next set of analyses, we test the hypothesis that successful recall of a naturalistic stimulus entails recovering the trajectory of a stimulus.

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/3_trajectory.pdf}
\caption{\small \textbf{Video and recall trajectory plots.} A). 2-dimensional embedding of the video events model (reduced with UMAP algorithm). The arrows indicate the forward direction of the video events. B). 2-dimensional embedding of the average recall events model.  The colors refer to the most similar video events. The directional lines connecting the points represent the true video event order. The arrows represent the group-average direction of all recall event transitions (i.e., a line segment connecting two consecutive recall event vectors) that intersected a circle (radius=.5) centered on the origin of the arrow.}
\label{fig:trajectory}
\end{figure}

To visualize the relationship between the video and recall event models, we embedded them into a 2D space using the UMAP dimensionality reduction algorithm ~\citep{McInHeal18}, where the points represent video/recall events and the distance between them represents their similarity/distance in ``topic space'' (Fig.~\ref{fig:trajectory}a).  We observed that visually, it appears that the two models have a very similar temporal evolution and geometric structure. To further quantify this correspondence and to characterize how participants navigated through the space during recall, we analyzed the angle between successive recall events. We created a grid of evenly spaced points (.25 units) in the 2D ``topic space''. For each point on the grid, we drew a circle (radius=.5) around the point and grouped together all recall transitions that intersected the circle (across subjects). To visualize the average angle, we converted each transition angle to a unit vector and then averaged the vectors together. To assess consistency in the direction of the recall transition across participants, we performed a Rayleigh test ($p<.001$, corrected at $p<.05$ using permutation procedure described in Methods). Notably, all video events fell within the path formed by the video events.  Thus, participants' recalls followed the same path as the video model (Fig.~\ref{fig:trajectory}b), suggesting the shape of the stimulus was preserved despite idiosyncratic differences between participants in their recalls.

@@

\subsubsection{Brain regions that mirror the correlational structure of naturalistic stimuli and memories}.
\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/4_brains_panels.pdf}
\caption{\small \textbf{Video and recall model correlational structure searchlight analysis.} A) The correlation matrix representing the temporal structure of the video model. Below it is a surface plot of brain regions whose activity patterns mirrored the correlational structure of video model during viewing.  B) The left matrix is S13's video-recall correlation matrix before warping. The yellow stars represent the path recovered by dynamic time warping. The right matrix is the result after warping. The same warping path was applied to the fMRI timeseries (separately for each participant). A surface plot of brain regions whose activity patterns mirrored the participant-specific recall correlation matrix. Significance threshold (p$<$.05) for both video and recall analyses derived from a permutation procedure by phase shifting the video model and recomputing the z-scores.}
\label{fig:brainz}
\end{figure}

Previous work suggests that the brain may represent our experiences as ``events'' with different time scales in a nested hierarchy ~\citep{LernEtal11, ChenEtal17, BaldEtal17}. In this next analysis we asked: what brain regions mirror the event-like correlational structure of our topic-based video model during viewing? We used a searchlight procedure (see Methods for details) to identify regions whose timepoint-by-timepoint correlational structure matched the correlational structure of our model. This analysis revealed a network of regions including bilateral frontal and cingulate cortex (Fig. ~\ref{fig:brainz}). Thus, these regions may play a role in maintaining information relevant to the narrative trajectory of the video.

While each participant remembered roughly the same sequence of events, there were subtle idiosyncracies in the way each individual described the events. For example, the event by event similarity structure (e.g. the off-diagonal values of the recall correlation matrices) differed across individuals (Supp. Fig.~\ref{fig:corrmats}). We hypothesized that idiosyncracies in the recall dynamics across subjects might be related to the temporal structure of brain networks are representing the events during video viewing. In this analysis, we identified brain areas whose responses at the time of viewing exhibited the same (segmented) idiosyncratic dynamics that later appeared in that participants’ recalls. Using a procedure similar to the approach above (see Methods for details), we searched for regions whose correlational structure during viewing was related to the idiosyncratic structure of the participant's recall model. The resulting map suggests that the correlational structure of a network of brain regions including the ventromedial prefrontal cortex, anterior cingulate, and right medial temporal lobe mirror the idiosyncratic structure of participants'' recalls (Fig. ~\ref{fig:brainz}).

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/5_topics.pdf}
\caption{\small \textbf{Analysis of topics by event memorability.} A). Group-averaged correlation between each video event and closest recall event. Error bars represent 95\% confidence intervals. Stars indicate top (green) and bottom (red) 3 remembered events across participants B). Wordles (top 200 words) representing a weighted average of video event topic vectors weighted by event memorability (e.g.\ the correlation values in A). The top wordle (green) contains words from the most remembered events and the bottom (red) contains words from least memorable events. The word sizes are proportional to the word's ``activation''. C). Trajectory represents the video event model embedded in a 2D space using UMAP. The large colored dots are video events. The small colored dots are all individual recall events across all participants, where the colors refer to the closest video event. The circular wordles represent the top 3 most (green circles) and least (red) memorable video events.  The left side of the wordle circles contain words associated with the video event vector and the right side contains words associated with the average recall event vector.}
\label{fig:topics}
\end{figure}

\subsubsection{Memorability by event and topics}
A notable advantage of representing naturalistic stimuli/memory using this approach is that the event models can be mapped back to the language used to fit the model(s).  This allows for analysis (and visualization) of the language used when participants' recounted their experiences.  In this next analysis, we 1) quantified the memorability of each video event and 2) used memorability to plot the top words for the most and least memorable scenes.

To measure video event memorability, we computed the correlation between each video event and the closest recall event and then averaged those values within participant (Fig.~\ref{fig:topics}a). We used event memorability to create a weighted average over all video event vectors, where video events with better memory were weighted more heavily. Then, we extracted the top 200 words and created ``wordles'' representing the most memorable themes in the video. The results reveal that scenes containing ``Sherlock'' and ``John'' were highly memorable, and also scenes containing words such as ``medium'', ``street'' and ``baker'' (Fig.~\ref{fig:topics}b). To find words related to the least memorable scenes, we inverted the memorability weights and again extracted the top 200 words.  Scenes containing ``Mike'' and ``Molly'' were least remembered, and also words like ``conference'', ``hospital'' and ``psychotherapist'' (Fig.~\ref{fig:topics}b).

Then, we took a closer look at the themes contained in the top/bottom 3 events, using the memorability weights to select the top/bottom events. We performed the same analysis as described above, but separately for the video and average recall event vectors representing each of the top/bottom 3 remembered events. In Figure~\ref{fig:topics}c, the trajectory represents a low-dimensional embedding of the video model and the large colored dots represent the video events. The smaller colored dots represent the distribution of individual recall events across all participants, colored by the closest video event. We plotted wordles for the top 3 (circled in green) and bottom 3 (red) most/least remembered events, where the video is represented in the left half of the circle and the average recall is represented in the right half of the circle. Visual inspection reveals that the words contained in the memorability weighted worldles (Fig.~\ref{fig:topics}b) overlap with the individual top/bottom event wordles (Fig.~\ref{fig:topics}c). These analyses shed light on the contents of naturalistic stimuli and accompanying memories, and highlight the flexibility in transforming between human and machine readable representations of the data afforded by our modeling approach.

\subsubsection{Additional measures of naturalistic recall}
Representing the video and verbal recall as events in ``topic space'' also affords us the ability to characterize the quality of recall in a more fine-grained and nuanced way than what was previously possible. We quantified the distance between each video event and its matching recall event (``precision''), the the distance between a video event and all non-matching recall events (``distinctiveness'') and the temporal order of the recall events (``order''). All three of these metrics were correlated to hand-annotated memory performance across participants ($p<.05$, see Supplemental Materials Section 1 for details).

Just like in a traditional ``free-recall'' list-learning experiment where participants view a list of words and then verbally recall them, our video-recall matching analysis approach affords us the ability to analyze memory in the same way. We considered 6 classic list-learning analyses: overall accuracy, probability of first recall \citep{Hoga75, HowaKaha99, Lami99}, lag-conditional response probability \citep{Kaha96, HowaKaha99}, serial position curve \citep{Murd62a}, temporal clustering and semantic clustering \citep{HowaKaha02, PolyEtal09}. Extending these classic list-learning analyses to naturalistic stimuli allows to directly compare trial- and naturalistic-based recall studies (see Supplemental Materials Section 2 for details).

@@

\subsubsection{Vector field analysis}
To quantify the flow of recall from event to event, we performed a vector field analysis.  We tiled the 2D topic space (x, y: -6 to 6 by .25) with an evenly spaced grid. For each grid point, we drew a circle around the point (radius=.5). Then, we tested whether any line segments (formed by event recall transitions) passed through this area of the topic space.  For example, say that a participant transitioned from recalling event 2 to event 3. These 2 recall events correspond to 2 points in topic space, and connecting them forms a line segment. We collected all line segments that passed through a given section of topic space (collapsing across participants). To plot the average direction of the line segments (i.e., the arrows for each grid point in ~\ref{fig:trajectory}b), we converted each of them to unit vectors and then averaged. For grid points where the direction was consistent (across all participants contributing to that point), the length of the arrow approaches 1, whereas if the direction was random the length of the arrow approaches 0. Lastly, we converted each unit vector to an angle (in radians) by taking the inverse tangent of the x, y components of the vector. To test whether the distribution of angles was significantly non-uniform (i.e., displayed a preferred direction across participants), we performed a Rayleigh test on the angles ($p<.001$, FDR-corrected at $p<.05$ using Benjamani-Hochberg procedure). Arrows where the Rayleigh test was significant are displayed in color while non-significant tests are displayed in gray with lower opacity.

@@

In the first analysis (Fig.~\ref{fig:topics}a/b), we quantified the most and least remembered topics/words throughout the entire video by computing a weighted average over all recall events, where the weights were proportional to memory for each recall event. To measure memory for each event, for each participant we computed the correlation between the video event vector and the closest recall event vector. We then averaged these correlation values across participants. We then computed a weighted average of all video events using the correlation values as weights. Next, we computed the dot product between this weighted-average video event vector and the R matrix (described in the paragraph above) to get ``activations'' for each word. Finally, we plotted the top 200 words where the size of the word is proportional to its ``activation''. To get the least remembered topics/words, we performed the same analysis but inverted memory weights.

In the second analysis (Fig.~\ref{fig:topics}c), we created wordles for the top/bottom 3 remembered video events indexed by the average correlation values (Fig.~\ref{fig:topics}a).  To get the ``activations'' for words associated with the video events, we computed the dot product between the video event vector and the R matrix. The same procedure was used to get word activations for the recall events. We then plotted the top 200 words for the top/bottom 3 recalled events.

>>>>

\begin{figure}[th!]
\centering
\includegraphics[width=1\textwidth]{figs/1_analysis_schematic_bw.pdf}
\caption{\small \textbf{Schematic of the analysis approach.} For each scene in the video, text descriptions were generated by hand. Three exemplary time points are displayed here.  Below the video descriptions are text samples from an example participant's verbal recall transcript.  We fit a topic model to the moment-by-moment video text descriptions and transformed participants' verbal recall transcripts using this same model. The bar charts display the resulting topic model weights for the video (in blue) and recall (in green) for three example topic dimensions.}
\label{fig:schematic}
\end{figure}

\subsubsection{Modeling the naturalistic stimulus and verbal recall}
We fit a topic model \citep{BleiEtal03} to hand-annotated text descriptions of scenes from the video. The text descriptions contained details of the scene such as the characters present, location, and a short summary of the scene (see Fig.~\ref{fig:schematic} for example text). As depicted in Fig.~\ref{fig:model}A, the topic vectors are sparse and change slowly over time. Furthermore, there are clear transitions from one topic ``state'' to the next, possibly indexing scene transitions in the stimulus. To get a better handle on this temporal structure, we computed a timepoints (1976) by timepoints (1976) correlation matrix of the video model (Fig.~\ref{fig:model}C).  This correlation matrix reveals that the model has a strong, block-diagonal structure with very low off-diagonal correlation values, suggesting the representations of each block are unique and highly descriminable. We found that the ``narrative details'' (a few sentences summarizing the dialogue and specific happenings of the scene) of the video were the most important feature driving this model structure (see Supp. Fig.~\ref{fig:feature-importance}A).

After watching the video, participants verbally recalled (in order) as much of the episode as they could.  We used the same topic model (fit with the text descriptions of the video) to transform participants' verbal recall transcripts. An example participant's (\#13) recall model is plotted in Fig.~\ref{fig:model}B. Notably, participant \#13's recall model appears visually similar to the video model. Similar to the video model analysis (Supp. Fig.~\ref{fig:feature-importance}A), we found that the ``narrative details'' feature was most important for driving the video/recall relationship (Supp. Fig.~\ref{fig:feature-importance}B).  Like the video model, topic vectors were sparse and changed gradually.  Then, we investigated the temporal structure of the recall matrices. From each participant's recall model, we computed a sentences (68 to 294) by sentences (68 to 294) correlation matrix (Fig.~\ref{fig:model}D). Like the video model, each participant's recall correlation matrix exhibited a strong block-diagonal structure (Supp. Fig.~\ref{fig:corrmats}). Notably, this suggests that participants recounted the video in discriminable segments, likely related to the recall of specific events from the video.

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{figs/2_eventseg.pdf}
\caption{\small \textbf{Modelling naturalistic stimuli and recall.} A depiction of our analysis pipeline. For all plots, darker colors indicate greater values and the range of each plot is 0-1.  \textbf{A.} A timepoints (1976) by topics (100) matrix representing the video stimulus.  Each row represents the most likely mixture of topics for a given timepoint (i.e., topic weights). Each column represents a different topic. \textbf{B.} A sentences (294) by topics (100) matrix representing participant \#13's recall. \textbf{C.} A viewing-time (1976) by viewing-time (1976) correlation matrix representing the correlation of each moment of the video model with every other moment of the video model. The yellow boxes represent `events' recovered by a hidden Markov model. \textbf{D.} A recall-time (294 sentences) by recall-time (294) correlation matrix for participant \#13. \textbf{E.} An events (34) by topics (100) matrix where each row represents the average topic vector for each event in the video model.  \textbf{F.} An events (27) by topics (100) matrix where each row represents the average topic vector for each event in participant \#13's recall model. \textbf{G.} A recall events (27) by video events (34) correlation matrix for participant \#13. The cells with a yellow border identify the video event with the highest correlation to a given recall event. \textbf{H.} A group averaged recall events (34) by video events (34) correlation matrix.  The cells with yellow borders are the video event with the highest correlation to a given average recall event.}
\label{fig:model}
\end{figure}

\subsubsection{Segmenting the video and recall models into ``events''}
As described above, a striking feature of both the video and recall correlation matrices is a strong, block structure along the diagonal of the matrices (see Fig.~\ref{fig:model}C,D).  We hypothesized that this structure might arise from transient stability in the "theme" of both the video and of participants' memory for the video. We hypothesized that specific events described by a participant could be ``matched'' to specific video events by computing the similarity (i.e., correlation) between their topic vectors. In other words, the video event that is most thematically similar to a particular recall event is the event that the participant is most likely describing. To test this idea, we segmented the video and recall models in $k$ events (i.e., states) using a hidden Markov model ~\citep{BaldEtal17}. Our algorithm determined 34 events for the video model and a range of values (range: 8-27; mean: 15.41; SD: 5.6) for the recall models (see Methods for details on choosing an optimal $k$ value).  The events discovered for the video model and a representative participant's recall model are highlighted as yellow rectangles outlining blocks along the diagonal of the correlation matrices (Fig.~\ref{fig:model}C,D).

Next, we created a video ``event model'' by averaging together neighboring topic vectors that the were classified to be in the same event, resulting in an events (34) by topics (100) matrix (Fig.~\ref{fig:model}E).  We performed the same procedure for the recall matrices (Fig.~\ref{fig:model}F). Then, we computed the correlation between video and recall event models, resulting in a video events (34) by recall events (8-27, depending on the participant) correlation matrix (Fig~\ref{fig:model}G for example, Supp. Fig~\ref{fig:matchmats} for all participants). These matrices represent the similarity (correlation) between each video event and each recall event (for each participant). To determine which video event a particular recall event was most likely describing, we found the index of the video event with the highest correlation to the recall event (i.e., the argmax).  This is depicted in Fig.~\ref{fig:model}G as the cells highlighted with a yellow border. Notably, our algorithm suggests that the example participant recalled most of the video events and did so in order.

Then, we computed a group-averaged recall event model and video-recall event correlation matrix. For each participant (and each recall event), we sorted the recall event vectors (across all participants) according to the video event with the highest correlation. We then averaged the recall event vectors within each group. This yielded an average recall event vector for all but one (of 34) video events, since no participant remembered one of the events according to our model. Lastly, we computed an average recall event (34) by video event (34) correlation matrix, and highlighted the cell with the highest correlation value with a yellow border (Fig.~\ref{fig:model}H). Notably, this matrix displayed high correlation values along the diagonal and low correlations in the off-diagonal cells. This suggests that on average, participants were able to recapitulate the events in the episode in a specific, sequential and highly descriminable way.

\subsubsection{The trajectory of a naturalistic experience is preserved in recall}
Classic approaches to studying free recall in episodic memory (such as overall accuracy and serial position curve) focus on the quantity of information recalled \citep{Murd62a}. More recent metrics (such as lag-CRP, semantic and temporal clustering) go further to describe list-level memory organization \citep{Kaha96, PolyEtal09}. However, these approaches cannot (and were not designed to) capture the rich temporal dynamics inherent in naturalistic stimuli and associated recall of those experiences. In the next set of analyses, we test the hypothesis that successful recall of a naturalistic stimulus entails recovering the trajectory of a stimulus.

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/3_trajectory.pdf}
\caption{\small \textbf{Video and recall trajectory plots.} \textbf{A.} 2-dimensional embedding of the video events model (reduced with UMAP algorithm). The arrows indicate the forward direction of the video events. \textbf{B.} 2-dimensional embedding of the average recall events model.  The colors refer to the most similar video events. The directional lines connecting the points represent the true video event order. The arrows represent the group-average direction of all recall event transitions (i.e., a line segment connecting two consecutive recall event vectors) that intersected a circle (radius=.5) centered on the origin of the arrow. \textbf{C.} Individual trajectory plots for each participant overlayed with the video model (black). The direction of the recall sequence is indicated by the direction of arrows connecting each recall event.  The colors refer to the most likely video event.}
\label{fig:trajectory}
\end{figure}

To visualize the relationship between the video and recall event models, we embedded them into a 2D space using the UMAP dimensionality reduction algorithm ~\citep{McInHeal18}, where the points represent video/recall events and the distance between them represents their similarity/distance in ``topic space'' (Fig.~\ref{fig:trajectory}A).  We observed that visually, it appears that the two models have a very similar temporal evolution and geometric structure. To further quantify this correspondence and to characterize how participants navigated through the space during recall, we analyzed the angle between successive recall events. We created a grid of evenly spaced points (.25 units) in the 2D ``topic space.'' For each point on the grid, we drew a circle (radius=.5) around the point and grouped together all recall transitions that intersected the circle (across subjects). To visualize the average angle, we converted each transition angle to a unit vector and then averaged the vectors together. To assess consistency in the direction of the recall transition across participants, we performed a Rayleigh test ($p<.001$, corrected at $p<.05$ using the permutation procedure described in Methods). Notably, the video events fell within the path formed by the recall events.  Thus, participants' recalls followed the same path as the video model (Fig.~\ref{fig:trajectory}B), suggesting the shape of the stimulus was preserved despite idiosyncratic differences between participants in their recalls.

@@

\subsubsection{Brain regions that mirror the correlational structure of naturalistic stimuli and memories}.
\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/4_brains_panels.pdf}
\caption{\small \textbf{Video and recall model correlational structure searchlight analysis.} \textbf{A.} The correlation matrix representing the temporal structure of the video model. Below it is a surface plot of brain regions whose activity patterns mirrored the correlational structure of video model during viewing.  \textbf{B.} The left matrix is S13's video-recall correlation matrix before warping. The yellow stars represent the path recovered by dynamic time warping. The right matrix is the result after warping. The same warping path was applied to the fMRI timeseries (separately for each participant). A surface plot of brain regions whose activity patterns mirrored the participant-specific recall correlation matrix. Significance threshold (p$<$.05) for both video and recall analyses derived from a permutation procedure by phase shifting the video model and recomputing the z-scores.}
\label{fig:brainz}
\end{figure}

Previous work suggests that the brain may represent our experiences as ``events'' with different time scales in a nested hierarchy ~\citep{LernEtal11, ChenEtal17, BaldEtal17}. In this next analysis we asked: what brain regions mirror the event-like correlational structure of our topic-based video model during viewing? We used a searchlight procedure (see Methods for details) to identify regions whose timepoint-by-timepoint correlational structure matched the correlational structure of our model. This analysis revealed a network of regions including bilateral frontal and cingulate cortex (Fig. ~\ref{fig:brainz}), suggesting that these regions may play a role in maintaining information relevant to the narrative trajectory of the video.

While each participant remembered roughly the same sequence of events, there were subtle idiosyncracies in the way each individual described the events. For example, the event-by-event similarity structure (i.e. the off-diagonal values of the recall correlation matrices) differed across individuals (Supp. Fig.~\ref{fig:corrmats}). We hypothesized that idiosyncracies in the recall dynamics across subjects might be related to the temporal structure of brain networks representing the events during video viewing. In this analysis, we identified brain areas whose responses at the time of viewing exhibited the same (segmented) idiosyncratic dynamics that later appeared in that participants’ recalls. Using a procedure similar to the approach above (see Methods for details), we searched for regions whose correlational structure during viewing was related to the idiosyncratic structure of the participant's recall model. The resulting map suggests that the correlational structure of a network of brain regions including the ventromedial prefrontal cortex, anterior cingulate, and right medial temporal lobe mirror the idiosyncratic structure of participants' recalls (Fig. ~\ref{fig:brainz}).

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/5_topics.pdf}
\caption{\small \textbf{Analysis of topics by event memorability.} \textbf{A.} Group-averaged correlation between each video event and closest recall event. Error bars represent 95\% confidence intervals. Stars indicate top (green) and bottom (red) 3 remembered events across participants \textbf{B.} Wordles (top 200 words) representing a weighted average of video event topic vectors weighted by event memorability (e.g.\ the correlation values in A). The top wordle (green) contains words from the most remembered events and the bottom (red) contains words from least memorable events. The word sizes are proportional to the word's ``activation''. \textbf{C.} Trajectory represents the video event model embedded in a 2D space using UMAP. The large colored dots are video events. The small colored dots are individual recall events across all participants, where the colors correspond to the closest video event. The circular wordles represent the top 3 most (green circles) and least (red) memorable video events.  The left side of the wordle circles contain words associated with the video event vector and the right side contains words associated with the average recall event vector.}
\label{fig:topics}
\end{figure}

\subsubsection{Memorability by event and topics}
A notable advantage of representing naturalistic stimuli and memory using this approach is that the event models can be mapped back to the language used to fit the model(s).  This allows for analysis (and visualization) of the language used when participants recounted their experiences.  In this next analysis, we 1) quantified the memorability of each video event and 2) used memorability to plot the ``top'' words for the most and least memorable scenes.

To measure video event memorability, we computed the correlation between each video event and the closest recall event and then averaged those values within participants (Fig.~\ref{fig:topics}A). We used event memorability to create a weighted average over all video event vectors, where video events with better memory were weighted more heavily. Then, we extracted the top 200 words and created ``wordles'' representing the most memorable themes in the video. The results reveal that scenes containing ``Sherlock'' and ``John'' were highly memorable, as were scenes containing words such as ``medium'', ``street'' and ``baker'' (Fig.~\ref{fig:topics}B). To find words related to the least memorable scenes, we inverted the memorability weights and again extracted the top 200 words.  Scenes containing ``Mike'' and ``Molly'' were least remembered, and also words like ``conference'', ``hospital'' and ``psychotherapist'' (Fig.~\ref{fig:topics}B).

Then, we took a closer look at the themes contained in the top/bottom 3 events, using the memorability weights to select the top/bottom events. We performed the same analysis as described above, but separately for the video and average recall event vectors representing each of the top/bottom 3 remembered events. In Figure~\ref{fig:topics}C, the trajectory represents a low-dimensional embedding of the video model and the large colored dots represent the video events. The smaller colored dots represent the distribution of individual recall events across all participants, colored by the closest video event. We plotted wordles for the top 3 (circled in green) and bottom 3 (red) most/least remembered events, where the video is represented in the left half of the circle and the average recall is represented in the right half of the circle. Visual inspection reveals that the words contained in the memorability weighted worldles (Fig.~\ref{fig:topics}B) overlap with the individual top/bottom event wordles (Fig.~\ref{fig:topics}C). These analyses shed light on the contents of naturalistic stimuli and accompanying memories, and highlight the flexibility in transforming between human and machine readable representations of the data afforded by our modeling approach.

\subsubsection{Additional measures of naturalistic recall}
Representing the video and verbal recall as events in ``topic space'' also affords us the ability to characterize the quality of recall in a more fine-grained and nuanced way than was previously possible. We quantified the distance between each video event and its matching recall event (``precision''), the the distance between a video event and all non-matching recall events (``distinctiveness'') and the temporal order of the recall events (``order''). All three of these metrics were correlated to hand-annotated memory performance across participants ($p<.05$, see Supplemental Materials Section 1 for details).

Just like in a traditional free recall list-learning experiment where participants view a list of words and then verbally recall them, our video-recall matching analysis approach affords us the ability to analyze memory in the same way. We considered 6 classic list-learning analyses: overall accuracy, probability of first recall \citep{Hoga75, HowaKaha99, Lami99}, lag-conditional response probability \citep{Kaha96, HowaKaha99}, serial position curve \citep{Murd62a}, temporal clustering, and semantic clustering \citep{HowaKaha02, PolyEtal09}. Extending these classic list-learning analyses to naturalistic stimuli allows to directly compare trial- and naturalistic-based recall studies (see Supplemental Materials Section 2 for details).

@@

\subsubsection{Vector field analysis}
To quantify the flow of recall from event to event, we performed a vector field analysis.  We tiled the 2D topic space (x, y: -6 to 6 by .25) with an evenly spaced grid. For each grid point, we drew a circle around the point (radius=.5). Then, we tested whether any line segments (formed by event recall transitions) passed through this area of the topic space.  For example, say that a participant transitioned from recalling event 2 to event 3. These 2 recall events correspond to 2 points in topic space, and connecting them forms a line segment. We collected all line segments that passed through a given section of topic space (collapsing across participants). To plot the average direction of the line segments (i.e., the arrows for each grid point in ~\ref{fig:trajectory}B), we converted each of them to unit vectors and then averaged. For grid points where the direction was consistent (across all participants contributing to that point), the length of the arrow approaches 1, whereas if the direction was random the length of the arrow approaches 0. Lastly, we converted each unit vector to an angle (in radians) by taking the inverse tangent of the x, y components of the vector. To test whether the distribution of angles was significantly non-uniform (i.e., displayed a preferred direction across participants), we performed a Rayleigh test on the angles ($p<.001$, FDR-corrected at $p<.05$ using Benjamani-Hochberg procedure). Arrows where the Rayleigh test was significant are displayed in color while non-significant tests are displayed in gray with lower opacity.

@@

In the first analysis (Fig.~\ref{fig:topics}A,B), we quantified the most and least remembered topics/words throughout the entire video by computing a weighted average over all recall events, where the weights were proportional to memory for each recall event. To measure memory for each event, for each participant we computed the correlation between the video event vector and the closest recall event vector. We then averaged these correlation values across participants. We then computed a weighted average of all video events using the correlation values as weights. Next, we computed the dot product between this weighted-average video event vector and the R matrix (described in the paragraph above) to get ``activations'' for each word. Finally, we plotted the top 200 words where the size of the word is proportional to its ``activation''. To get the least remembered topics/words, we performed the same analysis but inverted memory weights.

In the second analysis (Fig.~\ref{fig:topics}C), we created wordles for the top/bottom 3 remembered video events indexed by the average correlation values (Fig.~\ref{fig:topics}A).  To get the ``activations'' for words associated with the video events, we computed the dot product between the video event vector and the R matrix. The same procedure was used to get word activations for the recall events. We then plotted the top 200 words for the top/bottom 3 recalled events.
